---
title: "Pipeline for estimating perceptual scales with MLCM"
author: Guillermo Aguilar
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
```

# Imports

We first import functions that execute the analysis itself. These functions 
depends on the packages `MLCM` and `snow`. As shown below, the main function
to be used is called `estimate_scales'. Other imported functions are helper 
functions.

```{r sourcing function definitions}
source('mlcm_functions.R')
```


# Dataset

We need the path to the CSV file to be analysed.

```{r estimate additive model}
filename = '../data/example-data.csv'
```

The CSV file should be already in the format 'understood' by MLCM:

- every row is a trial
- the first column is called 'Resp', which contains the binary response to the trial.
- the four columns containing the stimuli indices presented in that trial. In
this case we use L1 and C1 to code the luminance and context of the first stimulus, 
respectively. Similarly L2 and C2 code the luminance and context of the second 
stimulus.
The entries are integers starting from 1. For the luminance dimension there is no
limit or restriction on the amount of values. However, the current implementation
allows only 2 contexts, i.e. the columns C1 and C2 can only contain the integers
1 or 2. 

# First analysis: model selection

## Logic
The first step in the analysis is to fit the two possible models that MLCM provides:
additive conjoint measurement (`add`) or the full, saturated model (`full`).

The additive model assumes that the difference between contexts
is additive. In our case, that means that the scale for context 2 is just a 
shifted version of the scale for context 1. 

The full model relaxes that assumption and allows the possiblity of more complex
interactions and not only a constant shift, at the expense of having more 
parameters to be found.

We fit both models to the same data, and check whether the full model is indeed
significantly better at explaining the data as the additive one.


## Estimates scales with additive model

We call the function `estimate_scales` to do the fit. The parameter `modeltype` 
controls which model we want to fit (`add` or `full`). 
We set the flag `do_bootstrap` to FALSE because for the moment we are only 
interested in the model fit and point estimate and
we do not want to calculate bootstrap confidence intervals yet.

```{r}
obs.add <- estimate_scales(filename, 
                           modeltype='add', 
                           do_bootstrap=FALSE)
```

We see the resulting scales values for the first context (column L), and the single additive
shift factor for the second context (column C).

## Estimate scales with full ('saturated') model

```{r estimate full model}
obs.full <- estimate_scales(filename, 
                            modeltype='full', 
                            do_bootstrap=FALSE)
```
Here we see that for each combination of L and C we obtain a scale value.

Note: It is normal that we obtain a warning ('fitted probabilities numerically 
0 or 1 occurred'). This only means that for some comparisons the observer was 
perfectly consistent (always chose the same answer). This is expected for easy 
trials.


## Model comparison  

We now compare the two models with each other. This is done using the likelihood 
ratio test (for which the statistic follows the $\chi^2$ distribution, Knoblauch & Maloney, 2012).

Note: the command for the test is called `anova`, but what we are doing here is 
*not* an analysis of variance (ANOVA).

```{r likelihood ratio test}
print(anova(obs.add, obs.full, test='Chisq'))
```

The test considers the reduction of the residual deviance with the full model, 
and evaluates if that reduction is signifficant. The column Pr(>Chi)
shows the p-value.

If the test is non-significant (p-value >= 0.05), that means the 
additive model is enough to explain the data. Then, on the next step 
you should choose the additive model (`modeltype='add'`).

If it is significant, then the 'full' model explains the data better than
the additive, and in the next step you should use the full model 
(`modeltype='full'`).



# Second analysis: goodness of fit, outlier removal and final scale estimates

After selection which model to use, we now are ready to estimate the final 
scales by evaluate the goodness of fit. The same function `estimate_scales` 
does this job when the flag `remove_outliers` is set to TRUE,

The function will fit the model and run the goodness of fit evaluation (GoF). 
Roughly speaking, the fitted model is used to simulate data coming from the
same underlying parameters (scales values) and the same amount of trials. 
Over many simulations we obtain the distribution of the model residuals,
and the range obtained over simulations is compared with the observed residuals. 
If the model assumptions hold, then the observed residuals should be in the range 
from the simulated residuals. If that is the case, then a p-value calculated
over the range will *not* be signifficant. In other words, 
the GoF evaluation 'passes' when the p-value >= 0.05.


Over the years we have observed that the GoF evaluation is very sensitive to
outliers, for example lapses. We have thus developed a procedure in which 
we remove trials with high residuals (one by one, starting from the 
highest) and re-evaluate GoF. We repeat this procedure iteratively until
the GoF test passes. Usually less than 20 trials (from a dataset of thousands) 
is needed to be removed in order for the GoF test to pass. 

This is what is implemented when the flag `remove_outliers` is set to TRUE,
and what we now do.

We also pass the flag `do_bootstrap` as TRUE so that the function calculates
bootstrap confidence intervals. The flag `normalized` is TRUE so that the final 
scales are normalized to 1. 

```{r goodness of fit evaluation and outlier removal}
estimate_scales(filename, 
                modeltype="full", 
                do_bootstrap=TRUE, 
                remove_outliers=TRUE, 
                normalized=TRUE,
                savecsv=TRUE,
                saverds=TRUE)
               
```


The final output file is a CSV file containing the final scale estimates.
This file ends with '...scales.csv'. The filename will contain the word
'add' or 'full', depending on the model, and can contain 'norm' when the scales 
are normalized to 1. 
If some trials needed to be removed, the filename will also contain the word 'or', 
which means "outliers removed".
In that case you will also find a CSV file with filename that finishes with 
`trialsremoved`. This file contains the trials that needed to be removed for the 
GoF test to pass. The first column in this file is the row number of that trial
on the original dataset file.

We also save two additional files (bootsamples and the Rds file) that are not
used in the following parts of the analysis pipeline but that are nonetheless 
saved for archiving purposes. 



