---
title: "Pipeline for estimating perceptual scales with MLCM"
author: "Guillermo Aguilar"
date: "`r Sys.Date()`"
output: pdf_document
params:
  participant: DEMO
  stimulus: whites
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
here::i_am("surround_brightness/analysis/analysis-mlcm.Rmd")
```

## Setup

The MLCM scale estimation is performed primarily by the `{mlcm}` package.
Here we have encapsulated the estimation in a higher level function `estimate_scales()`,
which also adds additional functionality.
This can be found in `mlcm_functions.R` here, which needs to be `source`d.
These functions depends on the packages `{mclm}` and `{snow}`.

(Here we also use the package `{here}` for more flexible and portable paths)

```{r sourcing function definitions}
library(here)
source(here(".", "surround_brightness", "analysis", "mlcm_functions.R"))
```


## Step 0: Dataset

We need the path to the CSV file to be analysed.
This gets constructed from participant initials, and stimulus name.

```{r dataset}
filename <- here(
  "surround_brightness", "data", "results", params$participant, "analyzed",
  paste(paste(params$participant, params$stimulus, sep = "_"), "csv", sep = ".")
)
```

The CSV file should be already in the format 'understood' by MLCM:

- every row is a trial
- the first column is called 'Resp', which contains the binary response to the trial.
- the four columns containing the stimuli indices presented in that trial.

In this case we use `I1` and `C1`
to code the intensity and context of the first stimulus, respectively.
Similarly `I2` and `C2` code the intensity and context of the second stimulus.
The entries are integers starting from `1`.

For the intensity dimension there is no limit or restriction on the amount of values.
However, the current implementation allows only 2 contexts,
i.e., the columns `C1` and `C2` can only contain the integers `1` or `2`.

## Step 1: model selection

### Logic
The first step in the analysis is to fit the two possible models that MLCM provides:
additive conjoint measurement (`add`)
or the full, saturated model (`full`).

The additive model assumes that the difference between contexts is additive.
In our case,
that means that the scale for context `2`
is just a shifted version of the scale for context `1`.

The full model relaxes that assumption
and allows the possiblity of more complex interactions beyond only a constant shift,
at the expense of having more parameters to be fit.

We fit both models to the same data,
and check whether the full model is indeed
significantly better at explaining the data as the additive one,
given that it has more free parameters and is thus expected to always fit better.


### Estimates scales with additive model

We call the function `estimate_scales` to do the fit.
The parameter `modeltype` controls which model we want to fit (`add` or `full`).
We set the flag `do_bootstrap` to `FALSE`
because for the moment we are only interested in the model fit and point estimate
and we do not want to calculate bootstrap confidence intervals yet.

```{r}
model_additive <- estimate_scales(filename,
  modeltype = "add"
)
```

We see the resulting scales values for the first context (column `I`),
and the single additive shift factor for the second context (column `C`).

### Estimate scales with `full` ('saturated') model

```{r estimate full model}
model_full <- estimate_scales(filename,
  modeltype = "full",
  savecsv = TRUE
)
```
Here we see that for each combination of `L` and `C` we obtain a scale value.

Note: It is normal that we obtain a warning
('fitted probabilities numerically 0 or 1 occurred').
This only means that for some comparisons
the observer was perfectly consistent (always chose the same answer).
This is expected for easy trials.

By passing `savecsv = TRUE`, we are also asking for
the resulting scales to be saved to`'<...>.scales.csv'`.
The filename will contain the word '`add`' or '`full`', depending on the model.


### Model comparison

We now compare the two models with each other.
This is done using the likelihood ratio test
(for which the statistic follows the $\chi^2$ distribution, Knoblauch & Maloney, 2012).

Note: the command for the test is called `anova`,
but what we are doing here is *not* an analysis of variance (ANOVA).

```{r likelihood ratio test}
ratio_test <- anova(model_additive, model_full, test = "Chisq")
print(ratio_test)
```

The test considers the reduction of the residual deviance with the full model,
and evaluates if that reduction is signifficant.
The column `Pr(>Chi)` shows the p-value.

If the test _is not_ significant (p-value >= 0.05),
that means the additive model is enough to explain the data.
Then, on the next step you should choose the additive model (`modeltype="add"`).

If it _is_ significant,
then the `full` model explains the data better than the additive,
and in the next step you should use the full model (`modeltype='full'`).

```{r select model}
alpha <- 0.05
modeltype <- ifelse(ratio_test$Pr[2] < alpha, "full", "add")
print(modeltype)
```


## Step 2: goodness of fit, outlier removal

After selection which model to use,
we now are ready to estimate the final scales by evaluate the goodness of fit.
The same function `estimate_scales` does this job when the flag `remove_outliers` is set to `TRUE`.

The function will fit the model and run the goodness of fit evaluation (GoF).
Roughly speaking,
the fitted model is used to simulate data
coming from the same underlying parameters (scales values)
and the same amount of trials.
Over many simulations we obtain the distribution of the model residuals,
and the range obtained over simulations is compared with the observed residuals.
If the model assumptions hold,
then the observed residuals should be in the range from the simulated residuals.
If that is the case, then a p-value calculated over the range will *not* be signifficant.
In other words,
the GoF evaluation 'passes' when the $p >= 0.05$.

Over the years we have observed that
the GoF evaluation is very sensitive to outliers, for example lapses.
We have thus developed a procedure in which we remove trials with high residuals
(one by one, starting from the highest) and re-evaluate GoF.
We repeat this procedure iteratively until the GoF test passes.
Usually less than 20 trials (from a dataset of thousands)
is needed to be removed in order for the GoF test to pass.

This is what is implemented when the flag `remove_outliers` is set to `TRUE`,
and what we now do.

```{r goodness of fit evaluation and outlier removal}
estimate_scales(filename,
  modeltype = modeltype,
  remove_outliers = TRUE,
  savecsv = TRUE
)
trimmed_data <- here(
  "surround_brightness", "data", "results", params$participant, "analyzed",
  paste(params$participant, params$stimulus, modeltype, "trimmed.csv", sep = "_")
)
```

If some trials needed to be removed,
the filename will also contain the word '`trimmed`', which means the outliers  were "trimmed" out.
In that case you will also find a CSV file with the filename ennding in `.outliers`.
This file contains the trials that needed to be removed for the  GoF test to pass.
The first column in this file is the row number of that trial on the original dataset file.

## Step 3: final scale estimates, with bootstrapped confidence intervals

We also pass the flag `do_bootstrap` as `TRUE`
so that the function calculates bootstrap confidence intervals.
The flag `normalized` is `TRUE` so that the final scales are normalized
such that the highest scale value is to 1.

```{r bootstrap}
estimate_scales(trimmed_data,
  modeltype = modeltype,
  remove_outliers = FALSE,
  do_bootstrap = TRUE,
  normalized = TRUE,
  savecsv = TRUE,
  saverds = TRUE,
)
```


The final output file is a CSV file containing the final scale estimates.
This file ends with `'<...>.scales.csv'`.
and can contain '`norm`' when the scales are normalized to 1.

We also save two additional files
-- `<...>.bootsamples.csv` and an `<...>.Rda` file.
These are used for achival purposes,
but not in the rest of the analysis pipeline.