---
title: "Pipeline for estimating perceptual scales with MLCM"
author: "Guillermo Aguilar, Joris Vincent"
date: "`r Sys.Date()`"
output: pdf_document
params:
  participant: DEMO
  stimulus: whites
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
```

## Overview

We estimate perceptual scales using Maximum-Likelihood Conjoint Measurement
(MLCM).
The `{mlcm}` package implements this estimation,
and some additional functions for diagnostics, etc.

For convenience, we have encapsulated this estimation in various functions
in `mlcm_functions.R` here, which needs to be `source`d.
These functions encapsulate the scale estimation,
but also parallelize the diagnostics and bootstrapping,
and provide some convenience functions for wrangling the in- and output.

These functions depends on the packages `{mclm}` and `{snow}`.
In the current notebook,
we also use the package `{here}` for more flexible and portable paths.

```{r source function definitions}
# Set project root directory
here::i_am("surround_brightness/analysis/analysis-mlcm.Rmd")
library(here)
source(here(".", "surround_brightness", "analysis", "mlcm_functions.R"))
```

The complete process of perceptual scale estimation that we use here
follows several steps:
0. Load the data
1. Select which MLCM model to fit
  1a. Fit the `full` or _saturated_ model,
      where all scale values are estimated
  1b. Fit the `add`itive model,
      where values of the two scales differ only by a single additive shift
  1c. Choose model by comparing the fitting quality of the two models
2. Evaluate Goodness-of-Fit of model
  2b. Remove putative outliers
3. Estimate 95% confidence of scale values,
   by bootstrapping confidence intervals
   
Lastly, we export and save final scale estimates and CI bounds
so that they can be used later.


## Step 0: Dataset

First, we load the data from which perceptual scales should be estimated.
We have this saved in a CSV file,
the name of which is constructed from participant initials, and stimulus name.

```{r path to data}
filename_data <- paste(paste(params$participant, params$stimulus, sep = "_"), "csv", sep = ".")
filepath_data <- here("surround_brightness", "data", "results", params$participant, "analyzed", filename_data)
```
The original data will be loaded from `r toString(filepath_data)`

The CSV file should be already in the format 'understood' by {MLCM}:

- each row is a single trial
- first column, called `Resp`, contains the (binary) response to the trial
- next (four) columns contain the stimuli indices presented in that trial

In this case we use `I1` and `C1`
to code the intensity and context of the first stimulus, respectively.
Similarly `I2` and `C2` code the intensity and context of the second stimulus.
The entries are integers starting from `1`.

For the intensity dimension there is no limit or restriction on the amount of values.
However, our current implementation allows only 2 contexts,
i.e., the columns `C1` and `C2` can only contain the integers `1` or `2`.

```{r load data}
trial_data <- read.csv(filepath_data, sep = ",")[c("Resp", "I1", "I2", "C1", "C2")]
print(paste("number of trials:", nrow(trial_data)))
print(head(trial_data))
```

## Step 1: Model selection

`{MLCM}` can fit two possible models:

- `add`itive conjoint measurement, 
  which assumes that the difference between contexts is additive.
  In our case,that means that the scale for context `2`
  is just a shifted version of the scale for context `1`.
- `full`, saturated model
  which relaxes that assumption
  and allows the possibility of more complex interactions
  beyond only a constant shift.

We want to fit the model that best captures our data.
We fit both models to the same data,
and check which model is better at explaining the data.


First, we fit the `full` model.
We call the main function `estimate_scales` to do the fit.
The parameter `modeltype` controls which model we want to fit (`add` or `full`).
```{r fit full model}
model_full <- estimate_scales(trial_data, modeltype = "full")
print(model_full)
```

> Note: It is normal that we obtain a warning
> ('fitted probabilities numerically 0 or 1 occurred').
> This just means that for some comparisons (trials)
> the response is perfectly consistent,
> i.e., the participant always chose the same answer.
> This is expected for easy trials.

Here we see that for each combination of `I` and `C` we obtain a scale value.


Now, we fit the `add`itive model.
```{r fit additive model}
model_additive <- estimate_scales(trial_data, modeltype = "add")
print(model_additive)
```
We see the resulting scales values for the first context (column `I`),
and the single additive shift factor for the second context (column `C`).


### Model comparison

We now compare the two models with each other.
However, the `full` model has more parameters to be fit,
which means it gets more degrees of freedom to improve the fit.
Thus, we want to compare the two fits,
while penalizing the `full` model for its extra degrees of freedom.
In other words, is the `full` model
_statistically significantly_ better at explaining the data
than the `add`itive one, given that it has more free parameters
(and is thus expected to always fit better).

This is done using the likelihood ratio test
(for which the statistic follows the $\chi^2$ distribution, Knoblauch & Maloney, 2012).

> Note: that we use the `anova` function for this test,
> but what we are doing here is *not* an analysis of variance (ANOVA).

```{r likelihood ratio test}
ratio_test <- anova(model_additive, model_full, test = "Chisq")
print(ratio_test)
```

The test considers the reduction of the residual deviance with the full model,
and evaluates if that reduction is statistically significant.
The column `Pr(>Chi)` shows the p-value.

We compare this p-value to an alpha level,
to binarize to statistical significance and thus select which model to use

```{r select model}
alpha <- 0.05
significance <- ratio_test$Pr[2] < alpha
modeltype <- ifelse(significance, "full", "add")
```

Here, the test `r paste0("_is", ifelse(significance, "", "not"), "_")` significant
(p = `r paste(ratio_test$Pr[2], ifelse(significance,"<",">="), alpha)`).
Thus, here we use the `r paste0("_", modeltype, "_")` model going forward.

```{r fit model}
model <- estimate_scales(trial_data, modeltype = modeltype)
print(model)
```


## Step 2: Evaluate Goodness-of-Fit (GoF)

Now that we know which of the model types (`full` or `add`) fits best,
let's see how good that fit exactly is.

One way to evaluate the Goodness-of-Fit (GoF) of a GLM like this,
is to look at the _residuals_:
the distances between each of the observed values (trial outcomes),
and what the fitted model would predicted for that trial.
```{r residuals}
residuals <- residuals(model$obj)
```
We could combine these residuals into a single value
such as a Root Mean Squared Error (RMSE)
```{r RMSE}
RMSE <- sqrt(mean(residuals^2))
print(RMSE)
```
But is this good, or bad?
In order to evaluate the GoF, we need to compare the residuals to something.

Instead of setting some criterion for the RMSE,
in MLCM we evaluate the robustness of the estimated parameters (scale values).
Roughly speaking,
the fitted model is used to _simulate_ data
using the same underlying parameters (scale values) 
and the same amount of trials.
Over many simulations we obtain a _distribution_ of model residuals
-- the distances between predicted values and actual (simulated) values.

We then compare the distribution obtained over simulations,
with the residuals from the actual (observed) data.
If the model assumptions hold,
then the _observed_ residuals should be
in the distribution from the _simulated_ residuals.
Thus, we calculate the p-value that the distribution of simulated residuals
is more extreme than the observed residuals.
In other words the Goodness-of-Fit *"passes"*
if this comparison is *not* statistically significant ($p >= 0.05$).

We use the diagnostics from {mlcm} for the GoF evaluation,
although our convenience wrapper can speed this up
by parallelizing across CPU cores (`ncores`).

```{r goodness-of-fit}
gof <- goodness_of_fit(model, ncores = 4)
print(gof$p)
```
Here, the GoF test is
`r paste("_", ifelse(gof$p < alpha, "", "not"), "significant_")`
(p = `r paste(gof$p, ifelse(gof$p < alpha, "<", ">="), alpha)`),
meaning that the observed residuals are
_`r ifelse(gof$p < alpha, "more", "less")`_ extreme than the simulated residuals,
and the fit is considered _`r ifelse(gof$p < alpha, "not", "")` good_.


### Removing outliers, to improve the fit
Over the years we have observed that
the GoF evaluation is very sensitive to _outliers_:
idiosyncratic trials where the observed response is very different
from what the model would predict.
Such outliers have a high residual after fitting,
which strongly influences the GoF.
These outlying trials do happen,
for example lapses where the participant wanted to indicate stimulus 1,
but misclicked and indicated stimulus 2.

We have thus developed a procedure in which we `remove_outliers()`.
Trials with high residuals are removed one by one, starting from the highest,
and GoF is re-evaluated after each removal.
We repeat this procedure iteratively, until the GoF test passes.
Usually less than 20 trials (from a dataset of thousands)
need to be removed in order for the GoF test to pass.

The function for this outlier-removal
takes in the observed data, and the initial model.
Again, this function can parallelize the evaluations
by specifying `ncores` to use.

```{r remove outliers}
removal <- remove_outliers(
  observed_data = trial_data,
  model = model,
  ncores = 4,
)
```

Here we see that in order to "pass" the Goodness-of-Fit test
(p = `r paste(removal$gof$p, ">=", alpha)`),
`r paste(nrow(removal$outliers))` had to be removed,
leaving a "trimmed" dataset of `r paste(nrow(removal$trimmed))` trials.

This outlier-removal is timeconsuming,
so we save the trimmed data to be used in the rest of the analysis
We save both the removed outliers (for archival purposes),
and the trimmed data to be used in the rest of the analysis,
with the filename containing the word `<...>.trimmed.csv`,

```{r save trimmed data}
filepath_trimmed <- here(
  "surround_brightness", "data", "results", params$participant, "analyzed",
  paste(paste(params$participant, params$stimulus, modeltype, sep = "_"), "trimmed", "csv", sep = ".")
)
write.csv(removal$trimmed_data, filepath_trimmed, quote = FALSE, row.names = FALSE)
```

For archival purposes, we also save the removed trials, i.e., the outliers,
to a CSV file with the filename ending in `<...>.outliers.csv`.
The first column in this file is
the row number of that trial on the original dataset file.

```{r save outliers}
filepath_outliers <- here(
  "surround_brightness", "data", "results", params$participant, "analyzed",
  paste(paste(params$participant, params$stimulus, modeltype, sep = "_"), "outliers", "csv", sep = ".")
)
write.csv(removal$outliers, filepath_outliers, quote = FALSE, row.names = FALSE)
```

Now, we re-estimate the final scale values on just the trimmed data,
and evaluate the GoF of that model.

```{r estimate trimmed scales}
model <- estimate_scales(removal$trimmed_data,
  modeltype = modeltype,
)
print(model)
gof <- goodness_of_fit(model, ncores = 4)
print(gof$p)
```


## Step 3: Bootstrapped confidence intervals

We also calculate bootstrapped confidence intervals.
We run many (e.g., 1000) simulations,
where on each simulation new trials are generated from the fitted model
and a model is fit to those simulated trials.
Then, from all the simulated model fits,
we calculate the intervals for each scale value,
that contains 95% of the simulations.
This is the confidence interval:
we are 95% confident that the True scale value falls within this interval.

Here, we also _normalize_ our scale values.
In the scale estimation, the absolute values for the scales
are not (directly) interpretable, they have no (clear) units.
Moreover, they can vary from simulation to simulation.
So, we normalize our scale values, such that the maximum is set to `1`.

```{r bootstrap}
bootstrap <- bootstrap_CIs(
  model, 
  normalized = TRUE, 
  nsim = 1000, 
  ncores = 4
)
```

This returns an object with the whole bootstrapping,
and includes the `CI_low` and `CI_high` bounds for each scale value.
```{r CI}
print(bootstrap$CI_low)
print(bootstrap$CI_high)
```

We also save all the bootstrap samples to a file `<...>.bootsamples.csv`
-- this is used for archival purposes, but not in the rest of the analysis pipeline.

```{r save bootstrap samples}
filepath_bootsamples <- here(
  "surround_brightness", "data", "results", params$participant, "analyzed",
  paste(paste(params$participant, params$stimulus, modeltype, "norm", sep = "_"), "bootsamples", "csv", sep = ".")
)
write.csv(bootstrap$samples, filepath_bootsamples, quote = FALSE, row.names = FALSE)
```

## Step 4: Export scales, CIs
Now that we have our final, best fitting scale value estimates,
and the 95% confidence intervals for these,
we extract all this into a single dataframe
-- and export that as a `<...>.scales.csv` file.
In the filename we also include the `modeltype`: `add` or `full`,
and `_norm_` when the scales are normalized to 1.

```{r extract scales}
# Extract scale values
df_scales <- extract_scales(model, bootstrap, normalized = TRUE)
print(df_scales)
```
```{r export scales to CSV}
# Export to CSV
filename_scales <- paste(paste(params$participant, params$stimulus, modeltype, "norm", sep = "_"), "scales", "csv", sep = ".")
filepath_scales <- here("surround_brightness", "data", "results", params$participant, "analyzed", filename_scales)
write.csv(df_scales, filepath_scales, quote = FALSE, row.names = FALSE)
```



### Archive

We also save an additional `<...>.Rda` file.
This is for archival purposes, but not used in the rest of the analysis pipeline.

```{r save to Rda}
filename_rda <- paste(paste(params$participant, params$stimulus, modeltype, "norm", sep = "_"), "Rda", sep = ".")
filepath_rda <- here("surround_brightness", "data", "results", params$participant, "analyzed", filename_rda)
save(params, model, bootstrap, gof, file = filepath_rda)
```
