---
title: "Pipeline for estimating perceptual scales with MLCM"
author: "Guillermo Aguilar, Joris Vincent"
date: "`r Sys.Date()`"
output: pdf_document
params:
  participant: DEMO
  stimulus: whites
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
```

## Overview

We estimate perceptual scales using Maximum-Likelihood Conjoint Measurement
(MLCM).
The `{mlcm}` package implements this estimation,
and some additional functions for diagnostics, etc.

For convenience, we have encapsulated this estimation in various functions
in `mlcm_functions.R` here, which needs to be `source`d.
These functions encapsulate the scale estimation,
but also parallelize the diagnostics and bootstrapping,
and provide some convenience functions for wrangling the in- and output.

These functions depends on the packages `{mclm}` and `{snow}`.
In the current notebook,
we also use the package `{here}` for more flexible and portable paths.

```{r source function definitions}
# Set project root directory
here::i_am("surround_brightness/analysis/analysis-mlcm.Rmd")
library(here)
source(here(".", "surround_brightness", "analysis", "mlcm_functions.R"))
```

The complete process of perceptual scale estimation that we use here
follows several steps:
0. Load the data
1. Select which MLCM model to fit
  1a. Fit the `full` or _saturated_ model,
      where all scale values are estimated
  1b. Fit the `add`itive model,
      where values of the two scales differ only by a single additive shift
  1c. Choose model by comparing the fitting quality of the two models
2. Evaluate Goodness-of-Fit of model
  2b. Remove putative outliers
3. Estimate 95% confidence of scale values,
   by bootstrapping confidence intervals
   
Lastly, we export and save final scale estimates and CI bounds
so that they can be used later.


## Step 0: Dataset

We need the path to the CSV file to be analysed.
This gets constructed from participant initials, and stimulus name.

```{r dataset}
filename <- here(
  "surround_brightness", "data", "results", params$participant, "analyzed",
  paste(paste(params$participant, params$stimulus, sep = "_"), "csv", sep = ".")
)
```

The CSV file should be already in the format 'understood' by MLCM:

- every row is a trial
- the first column is called 'Resp', which contains the binary response to the trial.
- the four columns containing the stimuli indices presented in that trial.

In this case we use `I1` and `C1`
to code the intensity and context of the first stimulus, respectively.
Similarly `I2` and `C2` code the intensity and context of the second stimulus.
The entries are integers starting from `1`.

For the intensity dimension there is no limit or restriction on the amount of values.
However, the current implementation allows only 2 contexts,
i.e., the columns `C1` and `C2` can only contain the integers `1` or `2`.



## Step 1: Model selection

`{MLCM}` can fit two possible models:

- `add`itive conjoint measurement, 
  which assumes that the difference between contexts is additive.
  In our case,that means that the scale for context `2`
  is just a shifted version of the scale for context `1`.
- `full`, saturated model
  which relaxes that assumption
  and allows the possibility of more complex interactions
  beyond only a constant shift.

We want to fit the model that best captures our data.
We fit both models to the same data,
and check which model is better at explaining the data.


First, we fit the `full` model.
We call the main function `estimate_scales` to do the fit.
The parameter `modeltype` controls which model we want to fit (`add` or `full`).
```{r fit full model}
model_full <- estimate_scales(filename, modeltype = "full")
```

> Note: It is normal that we obtain a warning
> ('fitted probabilities numerically 0 or 1 occurred').
> This just means that for some comparisons (trials)
> the response is perfectly consistent,
> i.e., the participant always chose the same answer.
> This is expected for easy trials.

Here we see that for each combination of `I` and `C` we obtain a scale value.


Now, we fit the `add`itive model.
```{r fit additive model}
model_additive <- estimate_scales(filename, modeltype = "add")
```
We see the resulting scales values for the first context (column `I`),
and the single additive shift factor for the second context (column `C`).


### Model comparison

We now compare the two models with each other.
However, the `full` model has more parameters to be fit,
which means it gets more degrees of freedom to improve the fit.
Thus, we want to compare the two fits,
while penalizing the `full` model for its extra degrees of freedom.
In other words, is the `full` model
_statistically significantly_ better at explaining the data
than the `add`itive one, given that it has more free parameters
(and is thus expected to always fit better).

This is done using the likelihood ratio test
(for which the statistic follows the $\chi^2$ distribution, Knoblauch & Maloney, 2012).

> Note: that we use the `anova` function for this test,
> but what we are doing here is *not* an analysis of variance (ANOVA).

```{r likelihood ratio test}
ratio_test <- anova(model_additive, model_full, test = "Chisq")
print(ratio_test)
```

The test considers the reduction of the residual deviance with the full model,
and evaluates if that reduction is statistically significant.
The column `Pr(>Chi)` shows the p-value.

We compare this p-value to an alpha level,
to binarize to statistical significance and thus select which model to use

```{r select model}
alpha <- 0.05
significance <- ratio_test$Pr[2] < alpha
modeltype <- ifelse(significance, "full", "add")
```

Here, the test `r paste0("_is", ifelse(significance, "", "not"), "_")` significant
(p = `r paste(ratio_test$Pr[2], ifelse(significance,"<",">="), alpha)`).
Thus, here we use the `r paste0("_", modeltype, "_")` model going forward.

```{r fit model}
model <- estimate_scales(trial_data, modeltype=modeltype)
```


## Step 2: goodness of fit, outlier removal

After selection which model to use,
we now are ready to estimate the final scales by evaluate the goodness of fit.
The same function `estimate_scales` does this job when the flag `remove_outliers` is set to `TRUE`.

The function will fit the model and run the goodness of fit evaluation (GoF).
Roughly speaking,
the fitted model is used to simulate data
coming from the same underlying parameters (scales values)
and the same amount of trials.
Over many simulations we obtain the distribution of the model residuals,
and the range obtained over simulations is compared with the observed residuals.
If the model assumptions hold,
then the observed residuals should be in the range from the simulated residuals.
If that is the case, then a p-value calculated over the range
will *not* be statistically significant.
In other words,
the GoF evaluation 'passes' when the $p >= 0.05$.

Over the years we have observed that
the GoF evaluation is very sensitive to outliers, for example lapses.
We have thus developed a procedure in which we remove trials with high residuals
(one by one, starting from the highest) and re-evaluate GoF.
We repeat this procedure iteratively until the GoF test passes.
Usually less than 20 trials (from a dataset of thousands)
is needed to be removed in order for the GoF test to pass.

This is what is implemented when the flag `remove_outliers` is set to `TRUE`,
and what we now do.

```{r goodness of fit evaluation and outlier removal}
estimate_scales(filename,
  modeltype = modeltype,
  remove_outliers = TRUE,
  savecsv = TRUE
)
trimmed_data <- here(
  "surround_brightness", "data", "results", params$participant, "analyzed",
  paste(params$participant, params$stimulus, modeltype, "trimmed.csv", sep = "_")
)
```

If some trials needed to be removed,
the filename will also contain the word '`trimmed`', which means the outliers  were "trimmed" out.
In that case you will also find a CSV file with the filename ennding in `.outliers`.
This file contains the trials that needed to be removed for the  GoF test to pass.
The first column in this file is the row number of that trial on the original dataset file.

## Step 3: final scale estimates, with bootstrapped confidence intervals

We also pass the flag `do_bootstrap` as `TRUE`
so that the function calculates bootstrap confidence intervals.
The flag `normalized` is `TRUE` so that the final scales are normalized
such that the highest scale value is to 1.

```{r bootstrap}
estimate_scales(trimmed_data,
  modeltype = modeltype,
  remove_outliers = FALSE,
  do_bootstrap = TRUE,
  normalized = TRUE,
  savecsv = TRUE,
  saverds = TRUE,
)
```


The final output file is a CSV file containing the final scale estimates.
This file ends with `'<...>.scales.csv'`.
and can contain '`norm`' when the scales are normalized to 1.

We also save two additional files
-- `<...>.bootsamples.csv` and an `<...>.Rda` file.
These are used for achival purposes,
but not in the rest of the analysis pipeline.
